Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-10-17 paper/sections/abstract.tex
[ paper ] minor fixies.

\begin{abstract}
In this paper, we describe our approach to reconstruct and verify in
the proof assistant for type theory \Agda, solutions obtained by
the  automatic theorem prover \Metis to problems in classical
propositional logic. The translation method consists of justifying
step-by-step each deduction found in the prover's solution. To carry
out this plan, we emulate the inference rules of the prover,
providing a version of them in the proof assistant. Finally, we
parse the output of the prover, build and analyze the respective
tree associated with the proof, for ending up with an \Agda code of
the proof for type-checking.

>>>>

\begin{abstract}
In this paper, we describe our approach to reconstruct and verify in
\Agda proof assistant, solutions generated by \Metis prover to problems in classical propositional logic. The translation method consists of justifying
step-by-step each deduction found in the prover's solution. To carry
out this plan, we emulate the inference rules of the prover,
providing a version of them in the proof assistant. Finally, we
parse the output of the prover, build and analyze the respective
tree associated with the proof, for ending up with an \Agda code of
the proof for type-checking.

### 2017-10-17 paper/sections/conclusions.tex
[ paper ] minor fixies.

We presented a reliable proof-reconstruction of \Metis solutions
for problems in classical propositional logic without logic.
The reconstruction approach in this study was designed to use
only a syntactic treatment of the logic without considering the
semantics. This aspect represent a difficulty but also an opportunity
to extend this work towards the support for first-order logic.
As a first future work, we see the support of \Metis propositional
fragment with equality.

Summarizing, we provided for each \Metis inference rule a theorem that
emulate the rule deduction. The emulated
version rules exposed in Section~\ref{ssec:emulating-inferences} admit
improvements, we left this for future work, particularly, a better
implementation of \simplify rule.
Some of the improvements are related with computational efficiency.
For instance, we use widely functions to check for syntactical equivalence
between propositions and to reorder propositions to match with another.
We believe both are expensive tasks and can get help in the analysis
step inside the translator tool.

@@

One of the main contribution of this study is
increasing the trustworthiness of these automatic provers,
in our case, \Metis. Justify a proof-object of a theorem prover has a real
significant impact for these automatic tools.
The reverse engineering task can reveal important issues or bugs
in many parts of these system (e.g., preprocessing, reasoning, or
deduction modules). While doing this research, we had the opportunity
to find some of those kind of bugs\footnote{Issues No.~2, No.~4. and
commit \name{8a3f11e} in the \Metis' official repository
\url{https://github.com/gilith/metis}}. Fortunately, all these bugs
were fixed quickly for release 20170810 by Joe Hurd.

>>>>

We presented a reliable proof-reconstruction of \Metis solutions
for problems in classical propositional logic without logic.
The reconstruction approach in this study was designed to use
only a syntactic treatment of the logic without considering the
semantics. This aspect represents a difficulty but also an opportunity
to extend this work towards the support of first-order logic.
As a future work, we propose extend the proof-reconstruction to
support the  proposition fragment with equality.

Summarizing, we provided for each \Metis inference rule a theorem that
emulate the rule deduction. The emulated
version rules exposed in Section~\ref{ssec:emulating-inferences} admit
improvements, we left this for future work, particularly, a better
coverage of the \simplify rule.
Some of the improvements are related with computational efficiency.
For instance, we use widely functions to check for syntactical equivalence
between propositions and to reorder propositions to match with another.
We believe both are expensive tasks and can get help in the analysis
step inside the translator tool.

@@

One of the main contribution of this study is
increasing the trustworthiness of these automatic provers,
in our case, \Metis. Justify a proof-object of a theorem prover has a real
significant impact for these automatic tools.
The reverse engineering task can reveal important issues or bugs
in many parts of these system (e.g., preprocessing, reasoning, or
deduction modules). While doing this research, we had the opportunity
to find some of those kind of bugs\footnote{Issues No.~2, No.~4. and
commit \name{8a3f11e} in the \Metis' official repository
\url{https://github.com/gilith/metis}}. Fortunately, all these bugs
were fixed quickly by Joe Hurd in 20170810 release .

### 2017-10-17 paper/sections/introduction.tex
[ paper ] minor fixies.

Many approaches have been proposed and some tools have been
implemented for proof-reconstruction in the last decades. These
programs are relevant not only because it helps to spread their
usage but they also increase the confidence of their users about
their algorithms and their correctness (see, for example, bugs in
\ATPs~\cite{Keller2013,Bohme2011,Fleury2014} and
\cite{Kanso2012}). We mention some of these tools in the related
work, Section \ref{sec:related-work}.

In this paper, we ported a subset of the \Metis' inference rules to
\Agda, the propositional fragmented without equality. The formalization of these rules allowed us to justify step-by-step \Metis' \TSTP derivations by translating automatically to \Agda by a proof-reconstruction tool built for this research named \Athena.
We structure the paper as follows.
We briefly mention some notation and syntax of the classical propositional logic in Section~\ref{ssec:CPL}.
In Section~\ref{ssec:metis-language-and-proofs}, we introduce the
\Metis prover and its deduction system.
In Section~\ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
We listed some conclusions and future work in
Section~\ref{sec:conclusions}. Some related works with this paper is presented in Section~\ref{sec:related-work}.
Finally, we provide a complete example in the appendix \textsc{A} of the proof reconstruction workflow.

>>>>

Many approaches have been proposed and some tools have been
implemented for proof-reconstruction in the last decades. These
programs are relevant not only because it helps to spread their
usage but they also increase the confidence of their users about
their algorithms and their correctness (see, for example, bugs in
\ATPs~\cite{Keller2013,Bohme2011,Fleury2014} and
\cite{Kanso2012}).
We mention some of these tools in Section~\ref{sec:related-work}.

In this paper, we ported a subset of the \Metis' inference rules to
\Agda, the propositional fragmented without equality. The formalization of
these rules allowed us to justify step-by-step \Metis' \TSTP derivations by
translating automatically to \Agda by a proof-reconstruction tool built for
this research named \Athena.
This paper has been organized in the following way.
The first section of the paper, Section~\ref{ssec:CPL}, will mention some
notation and syntax of the classical propositional logic.
In Section~\ref{ssec:metis-language-and-proofs}, we introduce the
\Metis prover and its deduction system.
In Section~\ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
Section~\ref{sec:conclusions} summarizes the main findings and includes a
discussion of future work.
Some related works with this paper is presented in Section~\ref{sec:related-work}.
Finally, we provide a small example in the appendix \textsc{A} of the
proof-reconstruction workflow.

### 2017-10-17 paper/sections/metis-language-and-proofs.tex
[ paper ] minor fixies.

\Metis is an automatic theorem prover written in \len{Standard ML}
for first-order logic with equality developed by John Hurd
\cite{hurd2003first}. This prover is suitable for proof
reconstruction since it provides well-documented proof-objects to
justify its deduction steps from the basis of only six inference
rules. There are already some tools that perform proof-reconstruction
with \Metis, some of these systems are the following.

\name{Isabelle/HOL} incorporated \Metis as a macro-step
reconstruction tool for explaining proof steps (usually \CNF goals)
replied from other \ATPs like \name{CVC4}, \name{Vampire}, or
\name{Z3} (more details in Ref.~\cite{paulson2007source}).
In \cite{Farber2015}, \Metis was used to provide a tactic for
\name{HOL Light}; this challenges others like \emph{MESON}
or the \emph{leanCoP} tactic that reconstructs proofs
delivered by \name{leanCoP}~\cite{Farber2016}.

@@

\Metis looks for a refutation to prove \verb!fof! problems.
A \verb!fof! problem contains the conjecture also named the \emph{goal} in the problem. This goal is stripped into subgoals by using the \strip rule. The subgoals are the conjuncts in a general conjunction that is equivalent to the aforementioned goal.
Therefore, to prove the goal, \Metis proves independently each subgoal since they tend to be less complex or even smaller propositions. At the end, the conjunction of all subgoal proofs must prove the goal. Nevertheless, this last reasoning is not explicit neither in the \TSTP derivations nor in the \Metis' documentation.
Since \Metis replies refutation proofs, to prove each subgoal, in the \TSTP derivation, the inference rule \negate introduces the negation of such subgoals. The \negate rule is a specification of the \emph{assume} rule.

@@

A proof-object delivered in a \Metis' proof encodes a natural
deduction proof. The deduction system uses six inference rules
\cite{hurd2003first} listed in Fig.~\ref{fig:metis-inferences}.
With these inference rules as the only valid deduction steps, \Metis
attempts to prove conjectures by refutation (see. e.g.,
\emph{falsium} in the last conclusion of the \TSTP derivation in
Fig.~\ref{fig:metis-example}).

>>>>

\Metis is an automatic theorem prover written in \len{Standard ML}
for first-order logic with equality developed by John Hurd
\cite{hurd2003first}. This prover is suitable for proof
reconstruction since it provides well-documented proof-objects to
justify its deduction steps from the basis of only six inference
rules. There are already some tools that perform proof-reconstruction
with \Metis. Some of these systems are the following.

\name{Isabelle/HOL} incorporated \Metis as a macro-step
reconstruction tool for explaining proof steps (usually \CNF goals)
replied from other \ATPs like \name{CVC4}, \name{Vampire}, or
\name{Z3} (more details in Ref.~\cite{paulson2007source}).
\citeauthor{Farber2015} in \cite{Farber2015} present tactic based on \Metis for
\name{HOL Light} capable to challenge others tactics like \emph{MESON}
or the \emph{leanCoP} tactic~\cite{Farber2016}.

@@

\Metis looks for a refutation to prove \verb!fof! problems.
A \verb!fof! problem contains the conjecture also named the \emph{goal} in the
problem. This goal is stripped into subgoals by using the \strip rule. The
subgoals are the conjuncts in a general conjunction that is equivalent to the
aforementioned goal.
Therefore, to prove the goal, \Metis proves independently each subgoal since
they tend to be less complex or even smaller propositions. At the end, the
conjunction of all subgoal proofs must prove the goal. Nevertheless, this last
reasoning is not explicit neither in the \TSTP derivations nor in the
documentation available about \Metis.
Since \Metis replies refutation proofs, to prove each subgoal, in the \TSTP
derivation, the inference rule \negate introduces the negation of such
subgoals. The \negate rule is a specification of the \emph{assume} rule.

@@

A proof-object delivered by \Metis encodes a natural
deduction proof. The deduction system uses six inference rules
\cite{hurd2003first} listed in Fig.~\ref{fig:metis-inferences}.
With these inference rules as the only valid deduction steps, \Metis
attempts to prove conjectures by refutation (see. e.g.,
\emph{falsium} in the last conclusion of the \TSTP derivation in
Fig.~\ref{fig:metis-example}).

### 2017-10-17 paper/sections/preliminaries.tex
[ paper ] reviewed to resolve section.
[ paper ] minor fixies.

\emph{Theorems.}
A sequent $Γ ⊢ φ$ represents a theorem, but we
define it as a relation between a set of propositions premises $Γ$,
and $φ$ as the sequent's conclusion.
The symbol $⊢$ is called turnstile.
If $Γ = ∅$, we write $⊢ φ$, and we say that $φ$ is a theorem.
Then, a theorem, $Γ ⊢ φ$ , means that there is a derivation with
conclusion $φ$ and with all (uncancelled) hypotheses in $Γ$.

@@

\begin{figure}
\[%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{}
\RightLabel{assume $φ$}
\UnaryInfC{$Γ , φ ⊢ φ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ$}
\RightLabel{weaken $ψ$}
\UnaryInfC{$Γ , ψ ⊢ φ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{}
\RightLabel{$⊤$-intro}
\UnaryInfC{$Γ ⊢ ⊤$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ ⊥$}
\RightLabel{$⊥$-elim $φ$}
\UnaryInfC{$Γ ⊢ φ$}
\end{bprooftree}
%}
\]

@@

\[%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ \wedge ψ$}
\RightLabel{$\wedge$-proj$_1$}
\UnaryInfC{$Γ⊢ φ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ \wedge ψ$}
\RightLabel{$\wedge$-proj$_2$}
\UnaryInfC{$Γ⊢ ψ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ$}
\RightLabel{$\vee$-intro$_1$ $ψ$}
\UnaryInfC{$Γ ⊢ φ \vee ψ$}
\end{bprooftree}
%}
\]

\[%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{$Γ ⊢ ψ$}
\RightLabel{$\vee$-intro$_2$ $φ$}
\UnaryInfC{$Γ ⊢ φ \vee ψ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ,φ ⊢ \gamma $}
\AxiomC{$Γ , ψ  ⊢ \gamma$}
\RightLabel{$\vee$-elim}
\BinaryInfC{$Γ , φ \vee ψ ⊢ \gamma$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ , φ ⊢ ψ$}
\RightLabel{$\Rightarrow$-intro}
\UnaryInfC{$Γ ⊢ φ \Rightarrow ψ$}
\end{bprooftree}
%}
\]

>>>>

\emph{Theorems.}
A sequent $Γ ⊢ φ$ represents a theorem, but we
define it as a relation between a set of propositions premises $Γ$,
and $φ$ as the conclusion of the sequent.
The symbol $⊢$ is called turnstile.
If $Γ = ∅$, we write $⊢ φ$, and we say that $φ$ is a theorem.
Then, a theorem, $Γ ⊢ φ$ , means that there is a derivation with
conclusion $φ$ and with all (uncanceled) hypotheses in $Γ$.

@@

\begin{figure}
\[%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{}
\RightLabel{assume}
\UnaryInfC{$Γ , φ ⊢ φ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ$}
\RightLabel{weaken}
\UnaryInfC{$Γ , ψ ⊢ φ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{}
\RightLabel{$⊤$-intro}
\UnaryInfC{$Γ ⊢ ⊤$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ ⊥$}
\RightLabel{$⊥$-elim}
\UnaryInfC{$Γ ⊢ φ$}
\end{bprooftree}
%}
\]

@@

\[%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ \wedge ψ$}
\RightLabel{$\wedge$-proj$_1$}
\UnaryInfC{$Γ⊢ φ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ \wedge ψ$}
\RightLabel{$\wedge$-proj$_2$}
\UnaryInfC{$Γ⊢ ψ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ$}
\RightLabel{$\vee$-intro$_1$}
\UnaryInfC{$Γ ⊢ φ \vee ψ$}
\end{bprooftree}
%}
\]

\[%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{$Γ ⊢ ψ$}
\RightLabel{$\vee$-intro$_2$}
\UnaryInfC{$Γ ⊢ φ \vee ψ$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ ,φ ⊢ \gamma $}
\AxiomC{$Γ , ψ  ⊢ \gamma$}
\RightLabel{$\vee$-elim}
\BinaryInfC{$Γ , φ \vee ψ ⊢ \gamma$}
\end{bprooftree}\quad
\begin{bprooftree}
\AxiomC{$Γ , φ ⊢ ψ$}
\RightLabel{$\Rightarrow$-intro}
\UnaryInfC{$Γ ⊢ φ \Rightarrow ψ$}
\end{bprooftree}
%}
\]
